---
title: "Assignment_1.rmd"
author: "Andrew Shenal"
date: "2025-09-04"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This cell loads the tidyverse package which is very helpful for data manipulation in R
```{r message=FALSE, warning=FALSE}

# Set up key functions for R and manipulating data
library(tidyverse)
```


The code below imports the data from the R project then uses two methods for basic descriptive statistics. Sapply (found in the R documentation) is used to acquire mean, median, and range variables. Summary() is used across the whole dataset as an alternative method of acquiring descriptive statistics. 

```{r warning=FALSE}
# Data acquired from Kaggle website: Berk Erisen - "Wind Turbine Scada Dataset: 2018 Scada Data of a Wind Turbine in Turkey".
Wind <- read_csv("./archive/T1.csv")


# Descriptive Statistics using sapply
means <- sapply(Wind, mean, na.rm = TRUE)
median <- sapply(Wind, median, na.rm = TRUE)
range <- sapply(Wind, range, na.rm = TRUE)

print(means)
print(median)
print(range)

# Descriptive Statistics using summary
summary(Wind)

```


I converted the wind speed variable which is in m/s to feet per second by multiplying the column by 3.28084. 

```{r}

# Variable Transformation
Wind %>% 
  mutate(`Wind Speed (feet/s)` = `Wind Speed (m/s)`*3.28084) %>% 
  head()

```

Below is a scatterplot containing wind speed on the x axis and power generated from the wind turbine on the y axis.

```{r}

plot(Wind$`Wind Speed (m/s)`, Wind$`LV ActivePower (kW)`,
     xlab = "Wind Speed (m/s)", ylab = "LV ActivePower (kW)")

```


Below is a histogram plot of the power created over the course of the data collection period. 

```{r}

hist(Wind$`LV ActivePower (kW)`,
     xlab = "LV ActivePower (kW)")

```
